Practical Optional Types for Clojure

### 2015-02-28 formal-model.tex
start of 3

%TODO EXPLAIN what is a proposition?
The first insight about occurrence typing is that
logical formulas
can be used to represent type information about our programs
by relating parts of the runtime environment to types
via propositional logic.
\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard logical connectives: implications, conjunctions,
disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
(\figref{main:figure:termsyntax}).

The particular part of the environment we reference in a
type proposition is called the \emph{object}.
Every expression has an object that is either \emph{empty}, written \emptyobject{}, 
which says 
this expression is not known to evaluate to a particular component
  of the current environment, or a 
variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
that exactly indicates how the value of this
expression can be derived from the current environment.
Type propositions can only reference non-empty objects.

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a set of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we use the proposition {\isprop{\NumberFull}{\x{}}}.

@@

Types include the top type, \emph{untagged} unions, functions, singleton types
and class instances. 
We abbreviate \Booleanlong{} as \Boolean{}, \Keywordlong{} as \Keyword{}
and \NumberFull{} and \Number{}.
The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
{\ValueTrue} as \True and {\ValueFalse} as {\False}.
Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
may refer to the function argument.
%Latent means they are relevant when the function is applied rather than evaluated.
They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

@@

The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
are used as assumptions in the then and else branch respectively.
If the result of the \ifliteral{} is true, then the value either
came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
which implies \thenprop{\prop{3}} is true. 
The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
similarly.
The T-Local rule connects the type system to the proof system over type propositions
via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
to derive a type for a variable.
Using this rule, the type system can then appeal to L-Update to refine the type
assigned to \x{}.

We are now equipped to type check 
\egref{example:conditionalflow}:
$$
\clj{(if (number? x) (inc x) 0)}
$$

The test expression uses T-App, 
$$
\judgement{\isprop{\UnionNilNum{}}{\x{}}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
since \numberhuh{} has type
{\ArrowOne{\x{}}{\Top}{\Boolean}
        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
      and \x{} has object \x{}.

Now we check both branches using the extended proposition environment as specified by T-If.
$$
\judgement{{\isprop{\UnionNilNum}{\x{}}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
$$
\judgement{{\isprop{\UnionNilNum}{\x{}}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
%\inc{} has type
%$$
%{\ArrowOne{\x{}}{\Number}{\Number}
%        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
%$$
%We can now check the conditional with T-If.
%$$
%\judgement{\isprop{\Number}{\x{}}}{\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\zeroliteral{}}}{\Number}}{\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}
%$$
%Finally the function can be checked with T-Abs
%$$
%\judgement{}{\hastype{\abs{\x{}}{\UnionNilNum}{\ ...}}
%                                             {\ArrowOne{\x{}}{\UnionNilNum}{\Number}
%        {\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
%  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$

>>>>

The first insight about occurrence typing is that
logical formulas
can be used to represent type information about our programs
by relating parts of the runtime environment to types
via propositional logic.
\emph{Type Propositions} \prop{} make assertions like ``variable \x{} is of type \NumberFull{}'' or
``variable \x{} is not \nil{}''---in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard logical connectives: implications, conjunctions,
disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
(\figref{main:figure:termsyntax}).

The particular part of the runtime environment we reference in a
type proposition is called the \emph{object}.
The typing judgement relates an object to every expression in the language.
An object is either \emph{empty}, written \emptyobject{}, 
which says 
this expression is not known to evaluate to a particular part
  of the current runtime environment, or a 
variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
that exactly indicates how the value of this
expression can be derived from the current runtime environment.
Type propositions can only reference non-empty objects.

The second insight is that we can replace the traditional 
representation of a
type environment (eg., a map from variables to types)
with a set of propositions, written \propenv{}. 
Instead of mapping \x{} to
the type \NumberFull{}, we use the proposition {\isprop{\NumberFull}{\x{}}}.

@@

Types include the top type, \emph{untagged} unions, functions, singleton types
and class instances. 
We abbreviate \Booleanlong{} as \Boolean{}, \Keywordlong{} as \Keyword{}
and \NumberFull{} and \Number{}.
The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
{\ValueTrue} as \True and {\ValueFalse} as {\False}.
Function types contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
may refer to the function argument.
%Latent means they are relevant when the function is applied rather than evaluated.
They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

@@

The propositions of the test expression \e{1}, \thenprop{\prop{1}} and \elseprop{\prop{1}}, are 
used as assumptions in the then and else branch respectively.
If the result of the \ifliteral{} is a true value, then it either
came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
which implies \thenprop{\prop{3}} is true. 
The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
similarly.
The T-Local rule connects the type system to the proof system over type propositions
via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
to derive a type for a variable.
Using this rule, the type system can then appeal to L-Update to refine the type
assigned to \x{}.

We are now equipped to type check
\egref{example:conditionalflow}, starting at body:
$$
\clj{... (if (number? x) (inc x) 0) ...}
$$

We know {\propenv{}} = {\isprop{\UnionNilNum{}}{\x{}}}.
The test expression uses T-App, 
$$
\judgement{\propenv{}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
since \numberhuh{} has type
{\ArrowOne{\x{}}{\Top}{\Boolean}
        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
      and \x{} has object \x{}.

Finally we check both branches using the extended proposition environment as specified by T-If.
Going down the then branch, our new assumption {\isprop{\Number}{\x{}}} is crucial to check
$$
\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\hastype{\x{}}{\Number{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
$$
because we can now satisify the premise of T-Local:
$$
\inpropenv{{\propenv{}},\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}.
$$
%\judgement{{\propenv{}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$
%$$
%\judgement{{\propenv{}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$

%\inc{} has type
%$$
%{\ArrowOne{\x{}}{\Number}{\Number}
%        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
%$$
%We can now check the conditional with T-If.
%$$
%\judgement{\isprop{\Number}{\x{}}}{\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\zeroliteral{}}}{\Number}}{\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}
%$$
%Finally the function can be checked with T-Abs
%$$
%\judgement{}{\hastype{\abs{\x{}}{\UnionNilNum}{\ ...}}
%                                             {\ArrowOne{\x{}}{\UnionNilNum}{\Number}
%        {\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
%  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$
